---
title: "Financial Tracking and Analysis with budgetoverview"
output: rmarkdown::html_vignette
description: >
  This vignette details how to use the budgetoverview package to manage financial data, create visualizations, and compare expenses to average spending.
vignette: >
  %\VignetteIndexEntry{Financial Tracking and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(budgetoverview)
```

# Adding Expenses

Begin by adding individual expenses using the add_expense function. This allows for tracking expenses by category.

```{r}
# Starting with an empty data frame for expenses
expenses <- data.frame(category = character(), amount = numeric())
# Adding a Food expense
expenses <- add_expense(expenses, "Food", 100)

```

# Financial Summary

Calculate your total expenses and how much you have saved with the calculate_financials function.

```{r}
# Example expenses data
expenses <- data.frame(category = c("Food", "Rent"), amount = c(100, 500))
# Calculating financials based on an income of 2000
financials <- calculate_financials(expenses, 2000)
```

# Expense Comparison

Compare your spending to the average with `compare_to_average`.

```{r}
# User's expenses data
user_expenses <- data.frame(category = c("Groceries", "Rent", "Utilities"),
                            amount = c(300, 1200, 150))
# Average Swiss expenses for comparison
swiss_averages <- c(Groceries = 350, Rent = 1000, Utilities = 200)
# Comparing and plotting
comparison_plot <- compare_to_average(user_expenses, swiss_averages)

```

# Visualizing Data

Visualize your financial data with `createBarChart` and `create_scatter_plot`.

```{r}
# Creating a bar chart of income, expenses, and savings
bar_chart <- createBarChart(test_data)
# Creating a scatter plot of expenses
scatter_plot <- create_scatter_plot(expenses)
```



