---
title: "Functions - How to use"
output: rmarkdown::html_vignette
description: >
  This vignette details how to use the budgetoverview package to manage financial data, create visualizations, and compare expenses to average spending.
vignette: >
  %\VignetteIndexEntry{Functions - How to use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(eval = FALSE)

library(budgetoverview)
```

# 1. add_expense

### Purpose

To add a new entry to a data frame that tracks expenses. It is designed to be interactive and reactive, fitting within the Shiny framework.

### Parameters

-   **`category`**: The category of the new expense. This is a string that describes what the expense is for (e.g., "Groceries", "Utilities").

-   **`amount`**: The monetary value of the expense. This should be a non-negative number. The function includes a check to ensure that the amount is not negative or **`NA`** (missing).

-   **`expenses_data`**: A reactive value that contains a data frame of expenses. In Shiny, reactive values are used to store and manage data that can change over time and trigger reactive updates in the app. This parameter expects the current state of expenses.

-   **`selected_categories`**: Another reactive value that stores a list of categories that have already been used or selected. This helps in tracking which expense categories have been entered.

-   **`session`**: The Shiny session object, used here for showing modal dialogues in the Shiny app.

### Functionality

1.  **Validate Amount**:

    -   The function first checks if the **`amount`** is **`NA`** (not available/missing) or less than 0. If either is true, a modal dialog box is displayed to the user, indicating that the amount must be non-negative, and the function returns **`FALSE`**.

2.  **Check Category and Update Data**:

    -   The function then checks if the **`category`** is not already present in **`selected_categories`**. If it's a new category, the function proceeds to add the expense.

    -   It appends the new expense data (category and amount) to the current **`expenses_data`** data frame using **`rbind`**.

    -   The **`expenses_data`** reactive value is updated with this new data frame.

    -   The **`selected_categories`** reactive value is also updated to include the new category.

    -   The function returns **`TRUE`** to indicate that the expense was successfully added.

3.  **Handle Duplicate Category**:

    -   If the category already exists in **`selected_categories`**, the function does not add the expense and returns **`FALSE`**.

### Example Usage in Shiny

In a Shiny app, this function would be used within an **`observeEvent`** or similar reactive context. It is triggered by a user action, specifically pressing an "Add Expense" button. The function reads the category and amount from the user input, attempts to add the expense, and then you can take additional actions based on whether the expense was successfully added.

```{r}
# Assuming you have a Shiny reactive environment set up

# Initialize reactive values
expenses_data <- reactiveVal(data.frame(category = character(), amount = numeric()))
selected_categories <- reactiveVal(character())

# Example usage within a Shiny observeEvent
observeEvent(input$add_expense, {
  result <- add_expense(input$category, input$expense, expenses_data, selected_categories)
  if (result) {
    # Perform actions after successful addition
    print("Expense added successfully.")
  } else {
    # Handle cases where addition fails
    print("Failed to add expense.")
  }
})
```

# 2. generate_bar_chart

Calculate your total expenses and how much you have saved with the calculate_financials function.

### Function Purpose

-   **Primary Use**: To visually represent financial data in the form of a bar chart. This kind of visualization makes it easier to interpret and compare different categories of financial data.

-   **Shiny Integration**: While it's designed for use in a Shiny application, it can also be used in other contexts where interactive visualizations are beneficial.

### Parameters

-   **`data`**: A data frame that contains the data to be plotted. This data frame is expected to have at least two columns:

    -   **`category`**: A column that contains the names of the categories (like "Income", "Expenses", "Savings").

    -   **`amount`**: A column that contains numerical values associated with each category.

-   **`colorblind_switch`**: A logical flag (TRUE or FALSE) that, when set to TRUE, enables a colorblind-friendly color palette for the bar chart. This is important for accessibility, ensuring that the visualization is readable by individuals with color vision deficiencies.

### Functionality

-   **Bar Chart Creation**:

    -   The function uses **`ggplot2`** to create a basic bar chart. The **`aes`** function specifies the aesthetics of the plot, mapping **`category`** to the x-axis and **`amount`** to the y-axis.

    -   **`geom_bar(stat = "identity")`** tells **`ggplot2`** to create a bar chart where the heights of the bars represent the **`amount`** values in the data.

-   **Color Palette**:

    -   Depending on the **`colorblind_switch`**, it selects a color palette that is either standard or colorblind-friendly. This is done using **`scale_fill_manual(values = color_palette)`**.

-   **Styling and Labels**:

    -   The chart is further styled with titles and labels using **`labs`**, and a minimal theme is applied for a clean, modern look (**`theme_minimal()`**). The legend is removed for simplicity.

-   **Plotly Conversion**:

    -   Finally, **`ggplotly(p)`** converts the **`ggplot`** object into a Plotly object. This step enhances the interactivity of the plot, allowing for features like hover effects in a web-based environment.

### Example Usage

-   In a Shiny application, this plot can be rendered using **`renderPlotly()`** to make it interactive.

```{r, eval = TRUE}
# Sample data
data <- data.frame(
  category = c("Income", "Expenses", "Savings"),
  amount = c(1000, 750, 250)
)

# Generate and display the bar chart
bar_chart <- generate_bar_chart(data, colorblind_switch = FALSE)
bar_chart
```

### Output

-   The function returns a Plotly ggplot object, which is an interactive plot that can be rendered in a web browser, making it suitable for Shiny applications or R Markdown documents.

# 3. generate_comparison_plot

### Function Purpose

-   **Primary Use**: The main goal of this function is to visualize and compare two sets of financial data -- the user's expenses and the average expenses in Switzerland -- in a scatter plot format. This helps in understanding how personal spending aligns with or differs from average spending patterns.

### Parameters

-   **`user_vs_swiss`**: A data frame that contains the data for comparison. This data frame should have three columns:

    -   **`category`**: Categories of expenses (like "Food", "Transport", "Utilities").

    -   **`user_amount`**: The amount of money spent by the user in each category.

    -   **`swiss_amount`**: The average amount spent in each category in Switzerland.

-   **`colorblind_switch`**: A logical flag that determines whether the plot should use a colorblind-friendly color palette. This enhances accessibility for users with color vision deficiencies.

### Functionality

-   **Data Visualization**:

    -   The function uses **`ggplot2`** to create a scatter plot. It plots both **`user_amount`** and **`swiss_amount`** as points on the plot, where each point's position is determined by its relative share in the total expenses and its absolute amount.

    -   The **`geom_point`** function is used twice to plot points for both user expenses and Swiss average expenses on the same graph, differentiated by shapes ("User" vs. "Swiss").

-   **Color Palette**:

    -   Depending on the **`colorblind_switch`**, it selects an appropriate color palette (**`brewer.pal`**) that is either standard or friendly for colorblind users.

-   **Styling and Labels**:

    -   The chart is styled with titles, labels, and minimal themes for a clean appearance. The legend is positioned to the right for clarity.

    -   The x-axis represents the percentage share of each category's expenses relative to the total, and the y-axis represents the actual amount spent.

-   **Plotly Conversion**:

    -   **`ggplotly(p)`** converts the **`ggplot`** object into a Plotly object, making the plot interactive. This interactivity enhances user engagement, allowing for features like tooltips on hover.

### Example Usage

-   Just like the previous ones, this function is intended to be used within a Shiny server function.

```{r}
# Sample data for user expenses vs. average expenses
user_vs_swiss <- data.frame(
  category = c("Food", "Transport", "Utilities"),
  user_amount = c(200, 150, 100),
  swiss_amount = c(250, 180, 120)
)

# Generate and display the comparison scatter plot
comparison_scatter_plot <- generate_comparison_plot(user_vs_swiss, colorblind_switch = FALSE)
comparison_scatter_plot
```

### Output

-   The function returns a Plotly ggplot object, which is an interactive plot suitable for web applications, particularly for Shiny.

# 4. generate_scatter_or_pie

### Purpose

Dynamically generate a visualization of expenses data, either as a scatter plot or a pie chart, based on user input.

-   **Adaptive Visualization**: The function can switch between two types of plots, offering flexibility in how the data is presented.

### Parameters

-   **`expenses_data_summary`**: A data frame containing summarized expense data. It should have at least two columns:

    -   **`category`**: The names of expense categories.

    -   **`percentage`**: The percentage share of each category in the total expenses.

    -   (The function seems to also use an **`amount`** column, as indicated in the scatter plot creation code.)

-   **`scatter_plot_type`**: A character string that specifies the type of plot to generate. It can be either "Scatter Plot" or "Pie Chart". This parameter allows the function to switch between plot types based on user preference.

-   **`colorblind_switch`**: A logical value that, when true, switches the color palette to a colorblind-friendly mode. This enhances accessibility for users with color vision deficiencies.

### Functionality

1.  **Color Palette Selection**:

    -   Based on the **`colorblind_switch`**, the function selects a suitable color palette using **`brewer.pal`**.

2.  **Creating the Scatter Plot**:

    -   If **`scatter_plot_type`** is "Scatter Plot", the function creates a scatter plot using **`ggplot2`**.

    -   The scatter plot represents each expense category as a point, with its position determined by the percentage share (x-axis) and the actual amount (y-axis).

    -   Additional features like point color, size, and labels are added for clarity and aesthetics.

    -   The scatter plot is then converted to an interactive Plotly plot using **`ggplotly`**.

3.  **Creating the Pie Chart**:

    -   If **`scatter_plot_type`** is "Pie Chart", the function uses Plotly's **`plot_ly`** to create a pie chart.

    -   The pie chart shows the proportion of each expense category in the total expenses.

    -   It includes features like labels, percentages, and hover information.

### Output

-   The function returns a Plotly object, which is an interactive plot. This interactivity enhances the user experience, particularly in web-based applications like Shiny.

### Example Usage in Shiny

-   The function can be called within a Shiny server function, where the user's choice of plot type and colorblind mode can be passed as inputs. The resulting plot can be rendered in the UI using **`renderPlotly`**.

```{r}
# Sample summarized expenses data
expenses_summary <- data.frame(
  category = c("Food", "Transport", "Utilities"),
  amount = c(200, 150, 100),
  percentage = c(40, 30, 30)
)

# Generate and display the scatter plot
scatter_plot <- generate_scatter_or_pie(expenses_summary, "Scatter Plot", colorblind_switch = FALSE)
scatter_plot

# Generate and display the pie chart
pie_chart <- generate_scatter_or_pie(expenses_summary, "Pie Chart", colorblind_switch = FALSE)
pie_chart
```
