---
title: "Installing budgetoverview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing budgetoverview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Installation Guide for budgetoverview

This vignette guides you through the installation process of the budgetoverview package.

# Installing from CRAN

If budgetoverview is available on CRAN, you can install it by running:

```{r}
install.packages("budgetoverview")
```

# Installing from GitHub

To install the development version from GitHub, use:

```{r}
# install.packages("devtools")  #Uncomment if devtools is not installed
devtools::install_github("your_username/budgetoverview")
```

Replace your_username with the actual GitHub username where the repository is hosted.

After installation, load budgetoverview with library(budgetoverview).

# Troubleshooting

If you encounter any issues, ensure your R and package dependencies are up to date. For specific errors, consult the documentation or the package's issue tracker on GitHub.

Replace `"your_username/budgetoverview"` with the actual path to your GitHub repository. This vignette provides instructions for both CRAN and GitHub installations, along with basic troubleshooting advice.
